---
title: "BSc thesis"
author: "Lana Broer"
output: pdf_document
---
Last updated: 13/06/2023


# Environment preparation 

First, we prepare the environment by clearing the global environment from any variables that may still be loaded. Then, we load all necessary packages.

```{r}
rm(list = ls())
```

## Load packages

```{r}
library(ADNIMERGE) # contains the data 
library(dplyr) # for data wrangling
library(corrplot) # to create correlation plot 
```


# Data preparation

In this subsection, we prepare the data for data cleaning. As such, we start by loading the data, after which we divide the original data frame into smaller subsets for clarity. 

## Loading data 

```{r}
data(adnimerge) # patient data 
data(datadic) # variable explanations
data(treatdis) # death records
```


# Data cleaning 

Before moving on to the descriptive statistics and data analysis, some data cleaning will be performed. This is mostly aimed at reducing the number of rows with missing diagnoses.

## Save original dataframe 

```{r}
original_adnimerge <- adnimerge
```

## Remove patients without baseline measurement 

```{r}
adnimerge <- adnimerge[!(is.na(adnimerge$DX) & adnimerge$VISCODE == "bl"),]
```

## Remove patients with only 1 measurement (censoring time = baseline)

```{r}
once <- table(adnimerge$RID) == 1
once <- names(once[once == T])

adnimerge <- adnimerge[!(adnimerge$RID %in% once),]

rm(once)
```

## Identify deaths 

```{r}
# identify deaths
death1 <- grepl('death', tolower(treatdis$WDREASON))
death2 <- grepl('death', tolower(treatdis$WDPARTCOM))
death3 <- grepl('die', tolower(treatdis$WDPARTCOM))
death4 <- grepl('expire', tolower(treatdis$WDREASON))
death5 <- grepl('expire', tolower(treatdis$WDPARTCOM))
treatdis$death <- death1 | death2 | death3 | death4 | death5

deaths <- treatdis[treatdis$death == T,]
deaths <- deaths[!(deaths$RID %in% c(222, 393, 5129, 1368, 2110, # manual check 
                                     2205, 4530, 4575, 4614, # showed these were 
                                     4652, 4888, 4947, 1260)), ] # not actual deaths

deaths_bool <- deaths[deaths$RID %in% unique(adnimerge$RID),] %>% 
  select(RID, EXAMDATE, death) %>%
  mutate(RID = as.numeric(RID),
         EXAMDATE = as.Date(EXAMDATE)) %>%
  arrange(RID, desc(EXAMDATE)) %>%
  filter(!duplicated(RID))

# add death data to all data 
baseline.dates <- adnimerge[adnimerge$VISCODE == "bl",] %>%
  mutate(RID = as.numeric(RID),
         EXAMDATE = as.Date(EXAMDATE)) %>%
  select(RID, EXAMDATE) %>%
  rename(bl.date = EXAMDATE)
deaths <- left_join(deaths_bool, baseline.dates, by = "RID") %>%
  mutate(Month.bl = as.numeric(difftime(EXAMDATE, bl.date, unit = 'days'))/(365.25/12),
         death = "Death",
         VISCODE = NA) %>%
  rename(DX = death) %>%
  select(RID, EXAMDATE, DX, VISCODE, Month.bl)

rm(baseline.dates, death1, death2, death3, death4, death5, deaths_bool)
```

## Convert data to long format 

To start with, data should be converted to long format with only the first and last date of each diagnosis available. 

```{r}
# correct long format containing only diagnoses and dates 
long <- adnimerge %>%
  select(RID, EXAMDATE, DX, VISCODE, EXAMDATE.bl, Month.bl) %>%
  mutate(Month.bl = round(Month.bl, 2),
         DX = as.character(DX))

# assume: if after dementia missing state, then stays in dementia 
for(i in 1:(nrow(long)-1)){
  if(long[i, "RID"] == long[i+1, "RID"]){
    if(!is.na(long[i, "DX"]) &
       is.na(long[i+1, "DX"]) &
       long[i, "DX"] == "Dementia")
      long[i+1, "DX"] = "Dementia"
  }
}

long <- long %>%
  arrange(RID, DX, EXAMDATE)

# remove in between measurements of the same diagnosis 
long$remove <- NA

for(i in 2:(nrow(long)-1)){
  if(long[i, "RID"] == long[i+1, "RID"] &
     !is.na(long[i, "DX"]) &
     !is.na(long[i+1, "DX"]) &
     long[i, "DX"] == long[i+1, "DX"]){
    
    if(long[i-1, "RID"] == long[i, "RID"]  &
       !is.na(long[i-1, "DX"]) &
       long[i-1, "DX"] == long[i, "DX"]){
      long[i, "remove"] = T
    }
    
  }
} 

long <- long[is.na(long$remove),] %>%
  arrange(RID, EXAMDATE)

# keep only the first date of diagnosis 
first_date <- long %>%
  filter(!is.na(DX)) %>%
  mutate(RID_DX = paste0(RID, "_", DX)) %>%
  arrange(RID, EXAMDATE) %>%
  filter(!duplicated(RID_DX)) %>%
  select(RID, DX, EXAMDATE) %>%
  rename(first = EXAMDATE)
first_date <- reshape(first_date, idvar = "RID", timevar = "DX", direction = "wide")

# keep only the last date of diagnosis
last_date <- long %>%
  filter(!is.na(DX)) %>%
  mutate(RID_DX = paste0(RID, "_", DX)) %>%
  arrange(RID, desc(EXAMDATE)) %>%
  filter(!duplicated(RID_DX)) %>%
  select(RID, DX, EXAMDATE) %>%
  rename(last = EXAMDATE)
last_date <- reshape(last_date, idvar = "RID", timevar = "DX", direction = "wide")

long <- left_join(long, first_date)
long <- left_join(long, last_date)
rm(first_date, last_date)

# remove patients that regress (MCI -> CN, dem -> MCI, dem -> CN) stages 
pt.before <- unique(long$RID)
long <- long %>%
  mutate(RID = ifelse((!is.na(first.MCI) & !is.na(last.CN) & first.MCI < last.CN) |
                        (!is.na(first.Dementia) & !is.na(last.CN) & first.Dementia < last.CN) |
                        (!is.na(first.Dementia) & !is.na(last.MCI) & first.Dementia < last.MCI), 
                      NA, RID))
pt.after <- unique(long$RID)[!is.na(unique(long$RID))]
length(setdiff(pt.before, pt.after))
rm(pt.before, pt.after)

long <- long[!is.na(long$RID),]

# remove NA values if they are in an interval between two of the same diagnoses
for(i in 1:(nrow(long))){
  if(is.na(long[i, "DX"])){
    if(!is.na(long[i, "first.CN"]) & 
       long[i, "first.CN"] <= long[i, "EXAMDATE"] &
       long[i, "EXAMDATE"] <= long[i, "last.CN"]){
      long[i, "RID"] = NA
    }else if(!is.na(long[i, "first.MCI"]) & 
       long[i, "first.MCI"] <= long[i, "EXAMDATE"] &
       long[i, "EXAMDATE"] <= long[i, "last.MCI"]){
      long[i, "RID"] = NA
    }else if(!is.na(long[i, "first.Dementia"]) & 
       long[i, "first.Dementia"] <= long[i, "EXAMDATE"] &
       long[i, "EXAMDATE"] <= long[i, "last.Dementia"]){
      long[i, "RID"] = NA
    }
  }
} 

long <- long[!is.na(long$RID), ] %>%
  arrange(RID, EXAMDATE) %>%
  select(RID, EXAMDATE, DX, VISCODE, Month.bl)

# keep only first and last NA value
for(i in 2:(nrow(long)-1)){
  if(long[i, "RID"] == long[i+1, "RID"]){
    if(is.na(long[i-1, "DX"]) & 
       is.na(long[i, "DX"]) &
       is.na(long[i+1, "DX"]))
      long[i, "RID"] = NA
  }
}
long <- long[!is.na(long$RID), ]

# add deaths 
deaths <- deaths[deaths$RID %in% long$RID,]
long <- rbind(long, deaths) %>%
  arrange(RID, EXAMDATE)
```


# Looking at transitions 

We are specifically interested in the transitions between states. In order to take a quick look at these transitions from a long format dataframe, we will write a function that produces a new dataframe with as output the patient ID, consecutive diagnoses and type of transitions.

```{r}
# function for crude look at transitions

transitions <- function(data){
  # keep only one line per diagnosis 
  for(i in 1:(nrow(data)-1)){
    if(data[i, "RID"] == data[i+1, "RID"]){
      if((!is.na(data[i, "DX"]) &
         !is.na(data[i+1, "DX"]) &
         data[i, "DX"] == data[i+1, "DX"]) | 
         (is.na(data[i, "DX"]) &
          is.na(data[i+1, "DX"]))){
        data[i, "RID"] = NA
      }
    }
  }
  data <- data[!is.na(data$RID), ] %>% 
    arrange(RID, Month.bl) %>%
    select(RID, DX, Month.bl)
  
  # add person indexing
  data$nr <- 1
  for(i in 1:(nrow(data)-1)){
    if(data[i, "RID"] == data[i+1, "RID"]){
      data[i+1, "nr"] = data[i, "nr"] + 1
    }
  }
  
  # name transitions 
  data$transitiontype <- as.character(NA)
  for(i in 1:(nrow(data)-1)){
    if(data[i, "RID"] == data[i+1, "RID"]){
      data[i, "transitiontype"] = paste0(data[i, "DX"], "_to_", data[i+1, "DX"])
    }
  }
  
  # add information about starting and final states
  # starting state
  ## unknown baseline diagnosis
  data$transitiontype <- ifelse(data$nr == 1 & is.na(data$transitiontype) 
                                 & is.na(data$DX),
                                 "NoTransitions_BaselineUnknown", data$transitiontype)
  ## known baseline diagnosis
  data$transitiontype <- ifelse(data$nr == 1 & is.na(data$transitiontype) 
                                 & !is.na(data$DX),
                                 paste0("NoTransitions_Baseline_", data$DX), data$transitiontype)
  
  # final state
  data$transitiontype <- ifelse(data$nr != 1 & is.na(data$transitiontype) 
                                 & is.na(data$DX),
                                 "FinalStateUnknown", data$transitiontype)
  
  data$transitiontype <- ifelse(data$nr != 1 & is.na(data$transitiontype) 
                                 & !is.na(data$DX),
                                paste0("Finalstate_", data$DX), data$transitiontype)
  
  data$nr <- NULL
  return(data)
}
```

## Transitions

```{r}
trans <- as.data.frame(table(transitions(long)$transitiontype))
```


# Adding states 

Since we are dealing with many NA values, we are introducing two extra states: 
- MCI-: a state that comes between CN and MCI
- MCI+: a state that comes between MCI and dementia

```{r}
# adding MCI- and MCI+
AddingStates <- left_join(long, transitions(long))
AddingStates$DX <- ifelse(is.na(AddingStates$DX) & AddingStates$transitiontype == "NA_to_MCI", "MCI-", AddingStates$DX)
AddingStates$DX <- ifelse(is.na(AddingStates$DX) & AddingStates$RID %in% c(520, 1190, 6001, 6327), "MCI-", AddingStates$DX)

for(i in 1:(nrow(AddingStates)-2)){
 if(AddingStates[i, "RID"] == AddingStates[i+1, "RID"] &
    !is.na(AddingStates[i, "transitiontype"]) &
    AddingStates[i, "transitiontype"] == "MCI_to_NA"){
   AddingStates[i+1, "DX"] = "MCI+"
   
   if(is.na(AddingStates[i+2, "DX"])) AddingStates[i+2, "DX"] = "MCI+"
 }
}

long2 <- AddingStates
long2$transitiontype <- NULL
rm(AddingStates)
```

## Censor patients that only have CN observations 

Since for patients with only CN observations, we cannot reasonably assume they will experience cognitive decline, we will be censoring these patients on their last known CN observation. 

```{r}
CN_to_NA_pat <- transitions(long2)
CN_to_NA_pat <- CN_to_NA_pat$RID[CN_to_NA_pat$transitiontype == "CN_to_NA"]
CN_to_NA <- long2[long2$RID %in% CN_to_NA_pat,]
CN_to_NA <- CN_to_NA[!is.na(CN_to_NA$DX),]

long3 <- long2[!(long2$RID %in% CN_to_NA_pat),]
long3 <- rbind(long3, CN_to_NA) %>%
  arrange(RID, EXAMDATE)

rm(CN_to_NA_pat, CN_to_NA)

# remove patients that only have one observation (baseline) left 
once <- table(long3$RID) == 1
once <- names(once[once == T])

long3 <- long3[!(long3$RID %in% once),]

rm(once)

# numbers per transition
trans3 <- as.data.frame(table(transitions(long3)$transitiontype))
```

## Take out MCI- state

Since patients have a probability of $1$ to leave the MCI- state to MCI, as this is how the MCI- state was defined, we create a new dataframe where we remove the MCI- states. 

```{r}
long4 <- long3[long3$DX != "MCI-",]

# numbers per transition
trans4 <- as.data.frame(table(transitions(long4)$transitiontype))

rm(long, long2, trans)
```

## Select medical variables 

We are mostly interested to see the effect of medical covariates on the transition probabilities. Therefore, we will be selecting which variables have enough measurements to be accounted for in the model. We will be taking the measurement that was taken at the first moment of diagnosis.

```{r}
# take out the baseline columns
names.df <- names(adnimerge)
names.df <- names.df[substr(names.df, nchar(names.df)-2, nchar(names.df)) != ".bl"]
takeout <- c("COLPROT", "ORIGPROT", "PTID", "SITE", "VISCODE",
             "AGE", "PTGENDER", "PTEDUCAT", "PTETHCAT", "PTRACCAT", "PTMARRY",
             "Month", "M", "DX", "FSVERSION", "IMAGEUID")
names.df <- setdiff(names.df, takeout)
adni_medical <- adnimerge[,names.df]

rm(names.df, takeout)

# look at the missingness (%) per column
missing <- as.data.frame(sapply(adni_medical, 
                                function(x) sum(is.na(x))/length(x)))

# add medical information to relevant follow-up moments
long4 <- long4 %>%
  mutate(RID = as.numeric(RID),
         EXAMDATE = as.Date(EXAMDATE))

adni_medical <- adni_medical %>%
  mutate(RID = as.numeric(RID),
         EXAMDATE = as.Date(EXAMDATE))

joined <- left_join(long4, adni_medical, by = c("RID", "EXAMDATE"))

# look at the missingness (%) per column
missing2 <- as.data.frame(sapply(joined[, 6:ncol(joined)], 
                                 function(x) sum(is.na(x))/length(x)))

# only values at first moment of diagnosis
adni_medical_first <- joined %>%
  arrange(RID, EXAMDATE) %>%
  mutate(interim = paste(RID, DX)) %>%
  filter(!duplicated(interim))

# look at the missingness (%) per column
missing3 <- as.data.frame(sapply(adni_medical_first[, 6:(ncol(adni_medical_first)-1)], 
                                 function(x) sum(is.na(x))/length(x)))  
names(missing3) <- "FreqMissing"
missing3$Var <- rownames(missing3)

rm(joined, missing, missing2)
```

### Consider only columns with enough values 

We will proceed only with the medical variables that have no more than $1/3$ of data missing.

```{r}
VarNames_EnoughInfo <- missing3$Var[missing3$FreqMissing < 1/3]
adni_medical_first2 <- adni_medical_first %>%
  select(RID, EXAMDATE, DX, VISCODE, Month.bl, all_of(VarNames_EnoughInfo))

rm(adni_medical_first, VarNames_EnoughInfo, missing3)
```


# Data preparation 

Since a lot of the measured medical variables are of similar nature, they will surely show colinearity. Therefore, we will be removing variables that are strongly correlated with (an)other variable(s). For this, we'll use a cut-off value of $0.65$ for the correlation coefficient. 

```{r}
# calculate correlations
cormatrix <- cor(adni_medical_first2[, 6:ncol(adni_medical_first2)], use = "pairwise.complete.obs")

# generate correlation plot 
pdf("corplot.pdf")
corrplot(cormatrix, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.6)
dev.off()

# extract pairs with > 0.65 correlation coefficient 
k <- 1 
pairsmatrix <- matrix(NA, nrow(cormatrix)*ncol(cormatrix), ncol = 2)

for(i in 1:nrow(cormatrix)){
  for(j in 1:ncol(cormatrix)){
    if(abs(cormatrix[i,j]) > 0.65){
      pairsmatrix[k,1] <- rownames(cormatrix)[i]
      pairsmatrix[k,2] <- colnames(cormatrix)[j]
      k <- k + 1 
    }
  }
}

pairsmatrix <- as.data.frame(pairsmatrix) %>%
  filter(!is.na(V1)) %>%
  filter(V1 != V2)

# keep only unique pairs 
for(i in 1:nrow(pairsmatrix)){
  if(pairsmatrix[i,1] > pairsmatrix[i,2]) pairsmatrix[i,] = pairsmatrix[i,c(2,1)]
}

pairsmatrix <- pairsmatrix %>%
  distinct(V1, V2)

# look at missingness in these variables 
v <- unique(c(pairsmatrix$V1, pairsmatrix$V2))
sort(apply(adni_medical_first2[, v], 2, function(x) sum(is.na(x))))

# remove variables that have high correlations (> 0.65)
med_variables <- colnames(adni_medical_first2[, 6:ncol(adni_medical_first2)])
med_variables <- setdiff(med_variables, c("ADAS11", "ADAS13", "ADASQ4", "CDRSB",
                                          "mPACCdigit", "mPACCtrailsB", "LDELTOTAL",
                                          "RAVLT.learning", "RAVLT.perc.forgetting",
                                          "WholeBrain", "Entorhinal", "MidTemp"))

# clear environment
rm(i,j, k, v, pairsmatrix, cormatrix)
```


# Final dataframe 

This dataframe contains only the variables we are interested in to possibly use for analysis. 

```{r}
# select non-medical final variables 
nonmed_variables <- c("RID", "DX", "Month.bl", 
                      "PTGENDER", "PTETHCAT", "PTRACCAT", 
                      "PTMARRY", "AGE", "PTEDUCAT")

# create final data frame 
adnimerge <- adnimerge %>%
  mutate(RID = as.numeric(RID),
         DX = as.character(DX))

adni_medical_first3 <- adni_medical_first2[,c("RID", "DX", 
                                              "Month.bl", med_variables)]

adni_final <- left_join(long4, adnimerge[,nonmed_variables])
adni_final <- left_join(adni_final, adni_medical_first3)
adni_final$VISCODE <- adni_final$EXAMDATE <- NULL

# dataframe for table 1 
baseline <- left_join(adni_final[adni_final$Month.bl == 0,], adnimerge) %>%
  select(colnames(adni_final), ADAS11, ADAS13, ADASQ4, CDRSB, mPACCdigit, 
         mPACCtrailsB, LDELTOTAL, RAVLT.learning, RAVLT.perc.forgetting,
         WholeBrain, Entorhinal, MidTemp)
```


# Clean environment

Save only the objects that are necessary for the files that follow hereafter.

```{r}
rm(adni_medical, adni_medical_first2, adni_medical_first3, adnimerge,
   deaths, long3, long4, trans3, trans4)
```
