---
title: "BSc thesis"
author: "Lana Broer"
output: pdf_document
---
Last updated: 13/06/2023


# Prepare environment 

## Load packages

```{r}
library(mstate) # to perform the analysis
library(colorspace) # colouring plots 
library(gtools) # categorize continuous variables
```


# Data preparation 

## Change months to years

```{r}
adni_final <- adni_final %>%
  rename(Year.bl = Month.bl) %>%
  mutate(Year.bl = Year.bl/12) 

rm(baseline)
```

## Transition matrix

```{r}
tmat_MCI <- transMat(x = list(c(2,3,4), 
                              c(3,4),
                              c(4),
                              c()),
                     names = c("MCI", "MCI+", "Dementia", "Death"))
```

## Categorizing numeric variables 

Because of convergence problems found in previous versions of the code, the decision was made to convert numerical variables to categorical. This will be done using the lower and upper $50\%$-quantiles, and is done in the chunk below.

```{r}
# save the original dataframe
adni_final_num <- adni_final

# categorize continuous variables
cont <- c("MMSE", "RAVLT.immediate", "RAVLT.forgetting", 
          "TRABSCOR", "FAQ", "Ventricles",
          "Hippocampus", "Fusiform", "ICV")

adni_final[,cont] <- apply(adni_final[,cont], 2, function(x) quantcut(x, q = 2, na.rm = TRUE))

adni_final <- adni_final %>%
  mutate(AGE = round(AGE),
         APOE4 = as.factor(APOE4))

# change quantiles to ascending factors
quant_values <- as.data.frame(apply(adni_final[,cont], 2, unique)) %>%
  filter(!is.na(MMSE))

for(i in colnames(quant_values)){
  if(i == "Ventricles" | i == "ICV" | i == "FAQ" | i == "TRABSCOR") quant_values[,i] = quant_values[c(2,1),i]
}

for(i in colnames(quant_values)){
  adni_final[,i] = factor(adni_final[,i], levels = quant_values[,i], 
                          labels = c("quant1","quant2"))
}

rm(cont)
```

## Data formatting to wide 

Before we can use the msprep function, we need the data to be in a suitable wide format.

```{r}
# save last follow-up time
maxtime <- adni_final %>% 
  select(RID, Year.bl) %>%
  arrange(RID, desc(Year.bl)) %>%
  filter(!duplicated(RID))

# start and stop for each DX 
startstop <- adni_final %>% 
  arrange(RID, Year.bl) %>%
  select(RID, DX, Year.bl)

for(i in 1:(nrow(startstop)-1)){
  if(startstop[i, "RID"] == startstop[i+1, "RID"] &
     startstop[i, "DX"] == startstop[i+1, "DX"]){
    startstop[i, "startstop"] = paste0(startstop[i, "DX"], ".start")
    startstop[i+1, "startstop"] = paste0(startstop[i+1, "DX"], ".stop")
  }
}

startstop$startstop <- ifelse(is.na(startstop$startstop), 
                              paste0(startstop$DX, ".start"), 
                              startstop$startstop)

startstop <- startstop %>% 
  arrange(RID, Year.bl) %>%
  select(RID, startstop, Year.bl) 

# mutate data to wide format 
wide <- reshape(startstop, idvar = "RID", timevar = "startstop", 
                      direction = "wide") %>%
  rename(CN.time = Year.bl.CN.start,
         MCI.time = Year.bl.MCI.start,
         MCIplus.time = `Year.bl.MCI+.start`,
         Dementia.time = Year.bl.Dementia.start,
         Death.time = Year.bl.Death.start) %>%
  mutate(CN.stat = ifelse(!is.na(CN.time), 1, 0),
         MCI.stat = ifelse(!is.na(MCI.time), 1, 0),
         MCIplus.stat = ifelse(!is.na(MCIplus.time), 1, 0),
         Dementia.stat = ifelse(!is.na(Dementia.time), 1, 0),
         Death.stat = ifelse(!is.na(Death.time), 1, 0)) %>%
  select(RID, CN.time, CN.stat, MCI.time, MCI.stat, MCIplus.time, MCIplus.stat,
         Dementia.time, Dementia.stat, Death.time, Death.stat)

wide <- left_join(wide, maxtime) %>%
  mutate(CN.time = ifelse(CN.stat == 0, Year.bl, CN.time),
         MCI.time = ifelse(MCI.stat == 0, Year.bl, MCI.time),
         MCIplus.time = ifelse(MCIplus.stat == 0, Year.bl, MCIplus.time),
         Dementia.time = ifelse(Dementia.stat == 0, Year.bl, Dementia.time),
         Death.time = ifelse(Death.stat == 0, Year.bl, Death.time), 
         
         CN.time = ifelse(CN.time == 0, NA, CN.time),
         MCI.time = ifelse(MCI.time == 0, NA, MCI.time),
         MCIplus.time = ifelse(MCIplus.time == 0, NA, MCIplus.time),
         Dementia.time = ifelse(Dementia.time == 0, NA, Dementia.time),
         Death.time = ifelse(Death.time == 0, NA, Death.time))
wide$Year.bl <- NULL

rm(maxtime, startstop)
```

Now, we need to add the transition specific covariate information. These variables are time-dependent, where at the first moment of (a new) diagnosis a new measurement was done for each variable. 

```{r}
# add transition specific covariates 
adni_final <- adni_final %>%
  arrange(RID, Year.bl)

# select variables that do not change throughout
meas_nonchanging <- adni_final[,c(1,4:10)] %>%
  arrange(RID, PTGENDER, APOE4) %>%
  filter(!duplicated(RID))

# select CN variables 
meas_CN <- adni_final[adni_final$DX == "CN",c(1, 11:19)] %>%
  filter(!duplicated(RID))
names(meas_CN)[-1] <- paste0(names(meas_CN)[-1], ".CN")

# select MCI variables
meas_MCI <- adni_final[adni_final$DX == "MCI",c(1, 11:19)] %>%
  filter(!duplicated(RID))
names(meas_MCI)[-1] <- paste0(names(meas_MCI)[-1], ".MCI")

# select MCI+ variables
meas_MCIplus <- adni_final[adni_final$DX == "MCI+",c(1, 11:19)] %>%
  filter(!duplicated(RID))
names(meas_MCIplus)[-1] <- paste0(names(meas_MCIplus)[-1], ".MCI+")

# select dementia variables
meas_dem <- adni_final[adni_final$DX == "Dementia",c(1, 11:19)] %>%
  filter(!duplicated(RID))
names(meas_dem)[-1] <- paste0(names(meas_dem)[-1], ".dem")
```

Next, we create dataframes for each diagnosis separately, and one dataframe that contains all measurements.

```{r}
# create baseline wide dataframe 
wide <- left_join(wide, meas_nonchanging)

# CN baseline dataframe
wide_CN <- left_join(wide, meas_CN)

# MCI baseline dataframe 
wide_MCI <- left_join(wide, meas_MCI)
wide_MCI$CN.time <- wide_MCI$CN.stat <- NULL

# dementia baseline dataframe
wide_dem <- left_join(wide, meas_dem)
wide_dem$CN.time <- wide_dem$CN.stat <- wide_dem$MCI.time <- wide_dem$MCI.stat  <- 
  wide_dem$MCIplus.time <- wide_dem$MCIplus.stat <- NULL

# add to the overall dataframe
wide_all <- left_join(wide, meas_CN)
wide_all <- left_join(wide_all, meas_MCI)
wide_all <- left_join(wide_all, meas_MCIplus)
wide_all <- left_join(wide_all, meas_dem)

rm(meas_CN, meas_MCI, meas_MCIplus, meas_dem, meas_nonchanging)
```

For the dataframes that contain only the CN and MCI measurements, we want to keep only the patients that had these diagnoses at baseline. So, for the `wide_MCI` dataframe, we keep only patients that had a baseline MCI diagnosis. 

```{r}
# split data on baseline diagnosis
baseline_MCI <- adni_final$RID[adni_final$Year.bl == 0 & adni_final$DX == "MCI"]
wide_MCI <- wide_MCI[wide_MCI$RID %in% baseline_MCI,]

rm(baseline_MCI, wide_dem, wide_CN, wide, wide_all)
```


# Model fitting 

We will be fitting the models that use only the patients with a MCI diagnosis at baseline. The models will be fit both on the study, and the age scale. Starting with the study scale.

## msprep 

```{r}
# covariates to be kept
covs.MCI <-  c("PTGENDER", "AGE", "APOE4", "MMSE.MCI", "RAVLT.immediate.MCI",
               "RAVLT.forgetting.MCI", "TRABSCOR.MCI", "FAQ.MCI", 
               "Ventricles.MCI", "Hippocampus.MCI", "Fusiform.MCI", "ICV.MCI")

# msprep for CN baseline 
msMCI <- msprep(data = wide_MCI,
                trans = tmat_MCI,
                time = c(NA, "MCIplus.time", "Dementia.time", "Death.time"),
                status = c(NA, "MCIplus.stat", "Dementia.stat", "Death.stat"),
                keep = covs.MCI)

# expanding the covariates
msMCI <- expand.covs(msMCI, covs.MCI)
```

## Fitting the model using msfit 

### Baseline MCI 

First, we fit a model with no covariates. 

```{r}
# msfit 
c0.MCI <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans),
               data = msMCI, method = "breslow")
msf0.MCI <- msfit(object = c0.MCI, vartype = "greenwood", trans = tmat_MCI)

# plot cumulative hazard 
plot(msf0.MCI, las = 1, 
     xlab = "Years since MCI diagnosis")

# probtrans
pt0.MCI <- probtrans(msf0.MCI, predt = 0, method = "greenwood")

# plot stacked transition probabilities 
pdf("MCI_studyscale_baseline.pdf")
print(plot(pt0.MCI, 
     ord = c(4,3,2,1),
     las = 1, 
     xlab = list("Years since inclusion with MCI", cex = 1),
     xlim = c(0,15), 
     type = "filled",
     use.ggplot = T,
     cols = c("red2", "orange", "light yellow",  "light green")))
dev.off()
```

Next, we fit the full model to determine the transition-specific covariate effects.

```{r, eval = T}
# fit full model on categorized data 
cfull.MCI <- coxph(Surv(Tstart, Tstop, status) ~ PTGENDERMale.1 + PTGENDERMale.2 +
                     PTGENDERMale.4 + PTGENDERMale.5 + PTGENDERMale.6 + 
                     AGE.1 + AGE.2 + AGE.3 + AGE.4 + AGE.6 + 
                     APOE41.1 + APOE41.2 + APOE41.4 + APOE41.6 + 
                     APOE42.1 + APOE42.2 + APOE42.4 + APOE42.6 +
                     MMSE.MCIquant2.1 + MMSE.MCIquant2.2 + MMSE.MCIquant2.3 + 
                     MMSE.MCIquant2.4 + MMSE.MCIquant2.6 + 
                     RAVLT.immediate.MCIquant2.1 + RAVLT.immediate.MCIquant2.2 + 
                     RAVLT.immediate.MCIquant2.4 + RAVLT.immediate.MCIquant2.6 + 
                     RAVLT.forgetting.MCIquant2.1 + RAVLT.forgetting.MCIquant2.2 + 
                     RAVLT.forgetting.MCIquant2.4 + RAVLT.forgetting.MCIquant2.6 + 
                     TRABSCOR.MCIquant2.1 + TRABSCOR.MCIquant2.2 + TRABSCOR.MCIquant2.4 +
                     FAQ.MCIquant2.1 + FAQ.MCIquant2.2 + FAQ.MCIquant2.4 +
                     Ventricles.MCIquant2.1 + Ventricles.MCIquant2.2 + Ventricles.MCIquant2.4 + 
                     Hippocampus.MCIquant2.1 + Hippocampus.MCIquant2.2 + Hippocampus.MCIquant2.4 + 
                     Fusiform.MCIquant2.1 + Fusiform.MCIquant2.2 + Fusiform.MCIquant2.4 +
                     ICV.MCIquant2.1 + ICV.MCIquant2.2 + 
                     strata(trans), 
                   data = msMCI, method = "breslow",
                   control = coxph.control(iter.max = 100))
```

```{r}
summary(cfull.MCI)
```

We need to check if the proportional hazards assumption is met, or that we have problems with non-proportionality. 

```{r}
ph.assump <- cox.zph(cfull.MCI)
ph.assump
# plot(ph.assump)
```

### Testing the Markov assumption 

Before moving on with this model, we want to check the Markov assumption. We will do that by adding the time of entering each state as covariates to the model. 

```{r, eval = F}
# make dataframe with entering times as covariates
wide_MCI_markov <- wide_MCI %>%
  mutate(MCIplus_entertime = ifelse(MCIplus.stat == 1, MCIplus.time, NA),
         Dementia_entertime = ifelse(Dementia.stat == 1, Dementia.time, NA))

# msprep with entering times as covriates
msMCI_markov <- msprep(data = wide_MCI_markov,
                       trans = tmat_MCI,
                       time = c(NA, "MCIplus.time", "Dementia.time", "Death.time"),
                       status = c(NA, "MCIplus.stat", "Dementia.stat", "Death.stat"),
                       keep = c(covs.MCI, "MCIplus_entertime", "Dementia_entertime"))

# expanding the covariates
msMCI_markov <- expand.covs(msMCI_markov, c(covs.MCI, "MCIplus_entertime", "Dementia_entertime"))

# fit model
cfull.MCI_markov <- coxph(Surv(Tstart, Tstop, status) ~ PTGENDERMale.1 + PTGENDERMale.2 +
                     PTGENDERMale.4 + PTGENDERMale.5 + PTGENDERMale.6 + 
                     AGE.1 + AGE.2 + AGE.3 + AGE.4 + AGE.6 + 
                     APOE41.1 + APOE41.2 + APOE41.4 + APOE41.6 + 
                     APOE42.1 + APOE42.2 + APOE42.4 + APOE42.6 +
                     MMSE.MCIquant2.1 + MMSE.MCIquant2.2 + MMSE.MCIquant2.3 + 
                     MMSE.MCIquant2.4 + MMSE.MCIquant2.6 + 
                     RAVLT.immediate.MCIquant2.1 + RAVLT.immediate.MCIquant2.2 + 
                     RAVLT.immediate.MCIquant2.4 + RAVLT.immediate.MCIquant2.6 + 
                     RAVLT.forgetting.MCIquant2.1 + RAVLT.forgetting.MCIquant2.2 + 
                     RAVLT.forgetting.MCIquant2.4 + RAVLT.forgetting.MCIquant2.6 + 
                     TRABSCOR.MCIquant2.1 + TRABSCOR.MCIquant2.2 + TRABSCOR.MCIquant2.4 +
                     FAQ.MCIquant2.1 + FAQ.MCIquant2.2 + FAQ.MCIquant2.4 +
                     Ventricles.MCIquant2.1 + Ventricles.MCIquant2.2 + Ventricles.MCIquant2.4 + 
                     Hippocampus.MCIquant2.1 + Hippocampus.MCIquant2.2 + Hippocampus.MCIquant2.4 + 
                     Fusiform.MCIquant2.1 + Fusiform.MCIquant2.2 + Fusiform.MCIquant2.4 +
                     ICV.MCIquant2.1 + ICV.MCIquant2.2 + 
                     MCIplus_entertime.4 + MCIplus_entertime.5 + 
                     Dementia_entertime.6 +
                     strata(trans), 
                   data = msMCI_markov, method = "breslow",
                   control = coxph.control(iter.max = 100))

# model outcomes 
summary(cfull.MCI_markov)

# clean environment
rm(wide_MCI_markov, msMCI_markov, cfull.MCI_markov)
```

### Probtrans for baseline MCI model

Next, we want to determine the transition probabilities for patients, given certain covariate values. We will make comparisons for several covariates:
- Ages: 60 vs. 70 vs. 80 
- Sex: male vs. female
- MMSE: quantile 1 vs. quantile 2 
- APOE4 allele: 0, 1, 2

```{r, eval = F}
# different ages 
x <- unique(msMCI[,c(1, 9:20)]) %>% 
  arrange(AGE) %>%
  group_by(PTGENDER, APOE4, MMSE.MCI, RAVLT.immediate.MCI, RAVLT.forgetting.MCI, 
           TRABSCOR.MCI, FAQ.MCI, Ventricles.MCI, Hippocampus.MCI,
           Fusiform.MCI, ICV.MCI) %>% 
  filter(n() > 1) %>%
  arrange(.by_group = T)

y <- unique(msMCI[msMCI$PTGENDER == "Male" & 
                   msMCI$APOE4 == 1 &
                   msMCI$MMSE.MCI == "quant2" & 
                   msMCI$RAVLT.immediate.MCI == "quant2" & 
                   msMCI$RAVLT.forgetting.MCI == "quant1" & 
                   msMCI$TRABSCOR.MCI == "quant1" & 
                   msMCI$FAQ.MCI == "quant1" & 
                   msMCI$Ventricles.MCI == "quant1" &
                   msMCI$Hippocampus.MCI == "quant2" &
                   msMCI$Fusiform.MCI == "quant2" &
                   msMCI$ICV.MCI == "quant2", 
                 c(1, 9:20)]) %>%
  filter(!is.na(id))

# choose c(717, 124) (60 vs. 72)

# different genders 
x <- unique(msMCI[,c(1, 9:20)]) %>% 
  group_by(AGE, APOE4, MMSE.MCI, RAVLT.immediate.MCI, RAVLT.forgetting.MCI, 
           TRABSCOR.MCI, FAQ.MCI, Ventricles.MCI, Hippocampus.MCI,
           Fusiform.MCI, ICV.MCI) %>% 
  filter(n() > 1) %>%
  arrange(.by_group = T)

y <- unique(msMCI[msMCI$AGE == 68 & 
                   msMCI$APOE4 == 0 &
                   msMCI$MMSE.MCI == "quant2" & 
                   msMCI$RAVLT.immediate.MCI == "quant2" & 
                   msMCI$RAVLT.forgetting.MCI == "quant1" &
                   msMCI$TRABSCOR.MCI == "quant1" &
                   msMCI$FAQ.MCI == "quant1" & 
                   msMCI$Ventricles.MCI == "quant1" &
                   msMCI$Hippocampus.MCI == "quant2" &
                   msMCI$Fusiform.MCI == "quant2" &
                   msMCI$ICV.MCI == "quant1", 
                 c(1, 9:20)]) %>%
  filter(!is.na(id))

# c(421, 604) (Male vs. Female)

# different MMSE quantiles 
x <- unique(msMCI[,c(1, 9:20)]) %>% 
  group_by(PTGENDER, AGE, APOE4, RAVLT.immediate.MCI, RAVLT.forgetting.MCI, 
           TRABSCOR.MCI, FAQ.MCI, Ventricles.MCI, Hippocampus.MCI,
           Fusiform.MCI, ICV.MCI) %>% 
  filter(n() > 1) %>%
  arrange(.by_group = T)

y <- unique(msMCI[msMCI$PTGENDER == "Female" & 
                   msMCI$AGE == 71 & 
                   msMCI$APOE4 == 0 &
                   msMCI$RAVLT.immediate.MCI == "quant2" & 
                   msMCI$RAVLT.forgetting.MCI == "quant2" & 
                   msMCI$TRABSCOR.MCI == "quant2" & 
                   msMCI$FAQ.MCI == "quant1" & 
                   msMCI$Ventricles.MCI == "quant1" &
                   msMCI$Hippocampus.MCI == "quant1" &
                   msMCI$Fusiform.MCI == "quant1" &
                   msMCI$ICV.MCI == "quant1",
                 c(1, 9:20)]) %>%
  filter(!is.na(id))

# c(779, 742) (quantile 1 vs. quantile 2)

# different APOE4 alleles
x <- unique(msMCI[,c(1, 9:20)]) %>% 
  arrange(APOE4) %>%
  group_by(PTGENDER, AGE, MMSE.MCI, RAVLT.immediate.MCI, RAVLT.forgetting.MCI, 
           TRABSCOR.MCI, FAQ.MCI, Ventricles.MCI, Hippocampus.MCI,
           Fusiform.MCI, ICV.MCI) %>% 
  filter(n() > 1) %>%
  arrange(.by_group = T)

y <- unique(msMCI[msMCI$PTGENDER == "Male" & 
                   msMCI$AGE == 71 & 
                   msMCI$MMSE.MCI == "quant2" &
                   msMCI$RAVLT.immediate.MCI == "quant2" & 
                   msMCI$RAVLT.forgetting.MCI == "quant1" & 
                   msMCI$TRABSCOR.MCI == "quant1" & 
                   msMCI$FAQ.MCI == "quant1" & 
                   msMCI$Ventricles.MCI == "quant1" &
                   msMCI$Hippocampus.MCI == "quant2" &
                   msMCI$Fusiform.MCI == "quant2" &
                   msMCI$ICV.MCI == "quant2",
                 c(1, 9:20)]) %>%
  filter(!is.na(id))

# choose c(401, 559) (0 vs. 1)

rm(x,y)
```

The study numbers for the 
- Ages: ID's 717, 124 (60 vs. 72)
- Sexes: ID's 421, 604 (male vs. female)
- MMSE quantiles: ID's 779, 742 (quantile 1 vs. quantile 2)
- APOE4 alleles: ID's 401, 559 (0 vs. 1)

Comparison between state occupation probabilities, for *different ages*:
ID's 717, 124 (60 vs. 72)

```{r}
age.ids <- c(717, 124)
pt.age.prob <- vector(mode = "list", length = length(age.ids))
pt.age.plots <- list()

for(i in 1:length(age.ids)){
  wh <- which(msMCI$id == age.ids[i])
  pt <- msMCI[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6
  
  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- "Female"
  pt[7,"APOE4"] <- 0
  
  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]
  
  pt$strata <- pt$trans
  msfage <- msfit(cfull.MCI, pt, trans = tmat_MCI)
  
  pt.age.prob[[i]] <- probtrans(msfage, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(age.ids)){
  pt.age.plots[[i]] <- plot(pt.age.prob[[i]], 
             main = list(paste0("Patient ", i, ": aged ", unique(msMCI$AGE[msMCI$id == age.ids[i]])), cex = 1.5),
             ord = c(4,3,2,1),
             las = 1, 
             xlab = list("Years since inclusion with MCI", cex = 1),
             xlim = c(0,15), 
             type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(wh, pt, msfage, age.ids)
```

Comparison between state occupation probabilities, for *different sexes*:
  Sexes: ID's 421, 604 (male vs. female)

```{r}
sex.ids <- c(421, 604)
sex <- c("Female", "Male")
pt.sex.prob <- vector(mode = "list", length = length(sex.ids))
pt.sex.plots <- list()

for(i in 1:length(sex.ids)){
  wh <- which(msMCI$id == sex.ids[i])
  pt <- msMCI[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6
  
  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- sex[i]
  pt[7,"APOE4"] <- 1

  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]

  pt$strata <- pt$trans
  msfsex <- msfit(cfull.MCI, pt, trans = tmat_MCI)

  pt.sex.prob[[i]] <- probtrans(msfsex, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(sex.ids)){
  pt.sex.plots[[i]] <- plot(pt.sex.prob[[i]], 
     main = list(paste0("Patient ", i, ": ", unique(msMCI$PTGENDER[msMCI$id == sex.ids[i]])), cex = 1.5),
     ord = c(4,3,2,1),
     las = 1, 
     xlab = list("Years since inclusion with MCI", cex = 1),
     xlim = c(0,15), 
     type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(sex, wh, pt, msfsex, sex.ids)
```

Comparison between state occupation probabilities, for *different MMSE quantiles*:
MMSE quantiles: ID's 779, 742 (quant1 vs. quant2)

```{r}
MMSE.ids <- c(779, 742)
MMSEquant <- c("[7,28]", "(28,30]")
pt.MMSE.prob <- vector(mode = "list", length = length(MMSE.ids))
pt.MMSE.plots <- list()

for(i in 1:length(MMSE.ids)){
  wh <- which(msMCI$id == MMSE.ids[i])
  pt <- msMCI[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6
  
  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- "Male"
  pt[7,"APOE4"] <- 0
  
  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]
  
  pt$strata <- pt$trans
  msfMMSE <- msfit(cfull.MCI, pt, trans = tmat_MCI)
  
  pt.MMSE.prob[[i]] <- probtrans(msfMMSE, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(MMSE.ids)){
  pt.MMSE.plots[[i]] <- plot(pt.MMSE.prob[[i]], 
       main = list(paste0("Patient ", i, ": MMSE in", MMSEquant[i]), cex = 1.5),
       ord = c(4,3,2,1),
       las = 1, 
       xlab = list("Years since inclusion with MCI", cex = 1),
       xlim = c(0,15), 
       type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(MMSEquant, wh, pt, msfMMSE, MMSE.ids)
```

Comparison between state occupation probabilities, for *different APOE4 alleles*:
APOE4 alleles: ID's 401, 559 (0 vs. 1)

Notice that there are actually 3 types of APOE4 alleles, however there are no patients with identical characteristics such that each allele occurs at least once. Therefore, we compare only allele 0 vs. 1, since these are the most prevalent in the dataset.

```{r}
APOE.ids <- c(401, 559)
apoes <- c(1,0)
pt.APOE.prob <- vector(mode = "list", length = length(APOE.ids))
pt.APOE.plots <- list()

for(i in 1:length(APOE.ids)){
  wh <- which(msMCI$id == APOE.ids[i])
  pt <- msMCI[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6

  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- "Female"
  pt[7,"APOE4"] <- apoes[i]

  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]

  pt$strata <- pt$trans
  msfAPOE <- msfit(cfull.MCI, pt, trans = tmat_MCI)

  pt.APOE.prob[[i]] <- probtrans(msfAPOE, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(APOE.ids)){
  pt.APOE.plots[[i]] <- plot(pt.APOE.prob[[i]], 
     main = list(paste0("Patient ", i, ": APOE allele", unique(msMCI$APOE4[msMCI$id == APOE.ids[i]])), cex = 1.5),
     ord = c(4,3,2,1),
     las = 1, 
     xlab = list("Years since inclusion with MCI", cex = 1),
     xlim = c(0,15), 
     type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(apoes, wh, pt, msfAPOE, APOE.ids)
```

## Fitting the model using msfit on the age-scale 

Next, we repeat all steps of the previous subsection on the age scale. 

### Baseline MCI 

```{r}
# create msfit object with shifted Tstart and Tstop
msMCI.age <- msMCI %>% 
  mutate(Tstart = Tstart + AGE,
         Tstop = Tstop + AGE)
```

First, we have the model without covariates. 

```{r}
# msfit 
c0.MCI.age <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans),
               data = msMCI.age, method = "breslow")
msf0.MCI.age <- msfit(object = c0.MCI.age, vartype = "greenwood", trans = tmat_MCI)

# plot cumulative hazard 
plot(msf0.MCI.age, las = 1, 
     xlab = "Years since MCI diagnosis")

# probtrans
pt0.MCI.age <- probtrans(msf0.MCI.age, predt = 0, method = "greenwood")

# plot stacked transition probabilities 
pdf("MCI_agescale_baseline.pdf")
print(plot(pt0.MCI.age, 
     ord = c(4,3,2,1),
     las = 1, 
     xlab = list("Years since inclusion with MCI", cex = 1),
     xlim = c(50,95), 
     type = "filled",
     use.ggplot = T,
     cols = c("red2", "orange", "light yellow",  "light green")))
dev.off()
```

Then, the model with all transition-specific covariate effects.

```{r}
# fit full model on categorized data 
cfull.MCI.age <- coxph(Surv(Tstart, Tstop, status) ~ PTGENDERMale.1 + PTGENDERMale.2 +
                     PTGENDERMale.4 + PTGENDERMale.5 + PTGENDERMale.6 + 
                     # AGE.1 + AGE.2 + AGE.3 + AGE.4 + AGE.6 + 
                     APOE41.1 + APOE41.2 + APOE41.4 + APOE41.6 + 
                     APOE42.1 + APOE42.2 + APOE42.4 + APOE42.6 +
                     MMSE.MCIquant2.1 + MMSE.MCIquant2.2 + MMSE.MCIquant2.3 + 
                     MMSE.MCIquant2.4 + MMSE.MCIquant2.6 + 
                     RAVLT.immediate.MCIquant2.1 + RAVLT.immediate.MCIquant2.2 + 
                     RAVLT.immediate.MCIquant2.4 + RAVLT.immediate.MCIquant2.6 + 
                     RAVLT.forgetting.MCIquant2.1 + RAVLT.forgetting.MCIquant2.2 + 
                     RAVLT.forgetting.MCIquant2.4 + RAVLT.forgetting.MCIquant2.6 + 
                     TRABSCOR.MCIquant2.1 + TRABSCOR.MCIquant2.2 + TRABSCOR.MCIquant2.4 +
                     FAQ.MCIquant2.1 + FAQ.MCIquant2.2 + FAQ.MCIquant2.4 +
                     Ventricles.MCIquant2.1 + Ventricles.MCIquant2.2 + Ventricles.MCIquant2.4 + 
                     Hippocampus.MCIquant2.1 + Hippocampus.MCIquant2.2 + Hippocampus.MCIquant2.4 + 
                     Fusiform.MCIquant2.1 + Fusiform.MCIquant2.2 + Fusiform.MCIquant2.4 +
                     ICV.MCIquant2.1 + ICV.MCIquant2.2 + 
                     strata(trans), 
                   data = msMCI.age, method = "breslow",
                   control = coxph.control(iter.max = 100))

# look at model outcomes
summary(cfull.MCI.age)
```

```{r}
# checking the proportional hazards assumption
ph.assump <- cox.zph(cfull.MCI.age)
ph.assump
# plot(ph.assump)
```

Comparison between state occupation probabilities, for *different ages*:
ID's 717, 124 (60 vs. 72)

```{r}
age.ids <- c(717, 124)
pt.age.prob_age <- vector(mode = "list", length = length(age.ids))
pt.age.plots_age <- list()
  
for(i in 1:length(age.ids)){
  wh <- which(msMCI.age$id == age.ids[i])
  pt <- msMCI.age[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6
  
  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- "Female"
  pt[7,"APOE4"] <- 0
  
  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]
  
  pt$strata <- pt$trans
  msfage <- msfit(cfull.MCI.age, pt, trans = tmat_MCI)
  
  pt.age.prob_age[[i]] <- probtrans(msfage, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(age.ids)){
  pt.age.plots_age[[i]] <- plot(pt.age.prob_age[[i]], 
             main = list(paste0("Patient ", i, ": aged ", unique(msMCI.age$AGE[msMCI.age$id == age.ids[i]])), cex = 1.5),
             ord = c(4,3,2,1),
             las = 1, 
             xlab = list("Age (years)", cex = 1),
             xlim = c(50,95), 
             type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(wh, pt, msfage, age.ids)
```

Comparison between state occupation probabilities, for *different sexes*:
  Sexes: ID's 421, 604 (male vs. female)

```{r}
sex.ids <- c(421, 604)
sex <- c("Female", "Male")
pt.sex.prob_age <- vector(mode = "list", length = length(sex.ids))
pt.sex.plots_age <- list()

for(i in 1:length(sex.ids)){
  wh <- which(msMCI.age$id == sex.ids[i])
  pt <- msMCI.age[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6
  
  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- sex[i]
  pt[7,"APOE4"] <- 1
  
  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]
  
  pt$strata <- pt$trans
  msfsex <- msfit(cfull.MCI.age, pt, trans = tmat_MCI)
  
  pt.sex.prob_age[[i]] <- probtrans(msfsex, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(sex.ids)){
  pt.sex.plots_age[[i]] <- plot(pt.sex.prob_age[[i]], 
       main = list(paste0("Patient ", i, ": ", unique(msMCI.age$PTGENDER[msMCI.age$id == sex.ids[i]])), cex = 1.5),
       ord = c(4,3,2,1),
       las = 1, 
       xlab = list("Age (years)", cex = 1),
       xlim = c(50,95), 
       type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(sex, wh, pt, msfsex, sex.ids)
```

Comparison between state occupation probabilities, for *different MMSE quantiles*:
  MMSE quantiles: ID's 779, 742 (quant1 vs. quant2)

```{r}
MMSE.ids <- c(779, 742)
MMSEquant <- c("[7,28]", "(28,30]")
pt.MMSE.prob_age <- vector(mode = "list", length = length(MMSE.ids))
pt.MMSE.plots_age <- list()

for(i in 1:length(MMSE.ids)){
  wh <- which(msMCI.age$id == MMSE.ids[i])
  pt <- msMCI.age[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6
  
  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- "Male"
  pt[7,"APOE4"] <- 0
  
  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]
  
  pt$strata <- pt$trans
  msfMMSE <- msfit(cfull.MCI.age, pt, trans = tmat_MCI)
  
  pt.MMSE.prob_age[[i]] <- probtrans(msfMMSE, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(MMSE.ids)){
  pt.MMSE.plots_age[[i]] <- plot(pt.MMSE.prob_age[[i]], 
       main = list(paste0("Patient ", i, ": MMSE in", MMSEquant[i]), cex = 1.5),
       ord = c(4,3,2,1),
       las = 1, 
       xlab = list("Age (years)", cex = 1),
       xlim = c(50,95), 
       type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(MMSEquant, wh, pt, msfMMSE, MMSE.ids)
```

Comparison between state occupation probabilities, for *different APOE4 alleles*:
APOE4 alleles: ID's 401, 559 (0 vs. 1)

Notice that there are actually 3 types of APOE4 alleles, however there are no patients with identical characteristics such that each allele occurs at least once. Therefore, we compare only allele 0 vs. 1, since these are the most prevalent in the dataset.

```{r}
APOE.ids <- c(401, 559)
apoes <- c(1,0)
pt.APOE.prob_age <- vector(mode = "list", length = length(APOE.ids))
pt.APOE.plots_age <- list()

for(i in 1:length(APOE.ids)){
  wh <- which(msMCI.age$id == APOE.ids[i])
  pt <- msMCI.age[rep(wh[1], 6), 9:20]
  pt$trans <- 1:6
  
  pt <- rbind(pt, pt[6,]) 
  pt[7,"PTGENDER"] <- "Female"
  pt[7,"APOE4"] <- apoes[i]
  
  attr(pt, "trans") <- tmat_MCI
  pt <- expand.covs(pt, covs.MCI)
  pt <- pt[1:6,]
  
  pt$strata <- pt$trans
  msfAPOE <- msfit(cfull.MCI.age, pt, trans = tmat_MCI)
  
  pt.APOE.prob_age[[i]] <- probtrans(msfAPOE, predt = 0)
}

# plot state occupation probabilities with ggplot
for(i in 1:length(APOE.ids)){
  pt.APOE.plots_age[[i]] <- plot(pt.APOE.prob_age[[i]], 
       main = list(paste0("Patient ", i, ": APOE allele", unique(msMCI.age$APOE4[msMCI.age$id == APOE.ids[i]])), cex = 1.5),
       ord = c(4,3,2,1),
       las = 1, 
       xlab = list("Age (years)", cex = 1),
       xlim = c(50,95), 
       type = "filled",
             use.ggplot = T,
             cols = c("red2", "orange", "light yellow",  "light green"))
}

rm(apoes, wh, pt, msfAPOE, APOE.ids)
```


# Saving all state occuption probability-plots to pdf's

## Study scale 

```{r, eval = F}
# age
for (i in 1:length(age.ids)) {
    file_name = paste("MCI_studyscale_age", unique(msMCI$AGE[msMCI$id == age.ids[i]]), ".pdf", sep="")
    pdf(file_name)
    print(pt.age.plots[[i]])
    dev.off()
}

# sex
for (i in 1:length(sex.ids)) {
    file_name = paste("MCI_studyscale_sex", unique(msMCI$PTGENDER[msMCI$id == sex.ids[i]]), ".pdf", sep="")
    pdf(file_name)
    print(pt.sex.plots[[i]])
    dev.off()
}

# MMSE
for (i in 1:length(MMSE.ids)) {
    file_name = paste("MCI_studyscale_MMSE", MMSEquant[i], ".pdf", sep="")
    pdf(file_name)
    print(pt.MMSE.plots[[i]])
    dev.off()
}

# APOE
for (i in 1:length(APOE.ids)) {
    file_name = paste("MCI_studyscale_APOE", unique(msMCI$APOE4[msMCI$id == APOE.ids[i]]), ".pdf", sep="")
    pdf(file_name)
    print(pt.APOE.plots[[i]])
    dev.off()
}
```

## Age scale 

```{r, eval = F}
# age
for (i in 1:length(age.ids)) {
    file_name = paste("MCI_agescale_age", unique(msMCI.age$AGE[msMCI.age$id == age.ids[i]]), ".pdf", sep="")
    pdf(file_name)
    print(pt.age.plots_age[[i]])
    dev.off()
}

# sex
for (i in 1:length(sex.ids)) {
    file_name = paste("MCI_agescale_sex", unique(msMCI.age$PTGENDER[msMCI.age$id == sex.ids[i]]), ".pdf", sep="")
    pdf(file_name)
    print(pt.sex.plots_age[[i]])
    dev.off()
}

# MMSE
for (i in 1:length(MMSE.ids)) {
    file_name = paste("MCI_agescale_MMSE", MMSEquant[i], ".pdf", sep="")
    pdf(file_name)
    print(pt.MMSE.plots_age[[i]])
    dev.off()
}

# APOE 
for (i in 1:length(APOE.ids)) {
    file_name = paste("MCI_agescale_APOE", unique(msMCI.age$APOE4[msMCI.age$id == APOE.ids[i]]), ".pdf", sep="")
    pdf(file_name)
    print(pt.APOE.plots_age[[i]])
    dev.off()
}
```
